Large and Infinitary Quotient Inductive-Inductive Types

### 2020-01-07 draft/paper.tex
type of refl
paper

\input{abbrevs.tex}
\newcommand{\Lift}{\Uparrow}
\newcommand{\ToS}{\mathsf{ToS}}
\newcommand{\ext}{\triangleright}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Pii}{\Pi}
\newcommand{\appi}{\mathsf{app}}
\newcommand{\lami}{\mathsf{lam}}
\newcommand{\Pie}{\Pi^{\mathsf{ext}}}
\newcommand{\appe}{\mathsf{app^{ext}}}
\newcommand{\lame}{\mathsf{lam^{ext}}}
\newcommand{\Piinf}{\Pi^{\mathsf{inf}}}
\newcommand{\appinf}{\mathsf{app^{inf}}}
\newcommand{\laminf}{\mathsf{lam^{inf}}}
\newcommand{\appitt}{\mathop{{\scriptstyle @}}}
\newcommand{\Refl}{\mathsf{Refl}}
\newcommand{\IdU}{\mathsf{IdU}}
\newcommand{\ReflU}{\mathsf{ReflU}}
\newcommand{\Sig}{\mathsf{Sig}}
\newcommand{\ToSSig}{\mathsf{ToSSig}}

@@

\todo{revise cETT (compress), more mention of interp of term constructors in model, include missing iso-cleaving to type subst}

@@

In Section \ref{sec:metatheory}, we describe a cumulative variant of extensional type
theory which is used as metatheory in the rest of the paper. In Section \ref{sec:tos},
we introduce the theory of QIIT signatures. TODO

@@

In this paper we consider QII algebras at arbitrary finite levels, along with
large eliminations, where the initial algebra can be at a different level than
the target algebra. For a simple example, consider natural number algebras at
level $i$, given as the $\Sigma$-type $NatAlg_i :\equiv(Nat : \Set_i)\times
Nat\times(Nat \ra Nat)$. The initial such algebra is the set of natural numbers,
which is at level $0$, but in type theory we often want to eliminate into larger
$\Set_i$, for example when computing a $Nat$-indexed family of types. To support
convenient reasoning about levels in this paper, we need two features:
cumulativity and a way to quantify over finite levels. The former is essential
to reduce bureaucratic overhead, while the latter is required to talk about
arbitrary levels.

\subsection{Cumulativity}

%% Cumulativity makes handling of algebras and constructions at different levels
%% much easier. For example, given an algebra $\gamma : NatAlg_i$, cumulativity
%% allows us to also have $\gamma : NatAlg_{i+j}$. In a non-cumulative setting such
%% as Agda, we may only have for each type $A : \Set_i$ a type $\mathsf{Lift}_{i\,j}\,A :
%% \Set_{i+j}$ which is isomorphic to $A$ up to definitional equality.

In its most basic form, cumulativity requires that whenever $A : \Set_i$, and $i
< j$ then also $A : \Set_j$. We also need cumulativity for $\Sigma$ and function
types, in order to have cumulativity for QII algebras in general. For example,
we want to have $\gamma : NatAlg_j$ whenever $\gamma : NatAlg_i$ and $i < j$.

We use Sterling's cumulative algebraic type theory \cite{sterling2019algebraic} as basis for our metatheory. \todo{REASONS, revamp}

%% The reason is that it is the only rigorous
%% algebraic definition of cumulativity that we are aware of, and it also

%% The reason for this is twofold. First, it
%% supports the right kind of cumulativity for our purposes. Second, it is itself a
%% finitary QIIT with sort equations, hence this paper provides a model theory for
%% it

We extend the base theory with an extensional identity type, $\Sigma$-types and
the unit type. In \cite{sterling2019algebraic} a proof of canonicity is provided for the base
theory, which also includes a standard set-theoretic model. It is
straightforward to extend the canonicity proof to cover our additional type
formers.

From now on, we refer to this theory as cETT (cumulative extensional type
theory). When working in cETT, we use the following notation. We have
Russell-style universes $\Set_i$ indexed by natural numbers, dependent functions
as $(x : A)\ra B$, and dependent pairs as $(x : A)\times B$ with projections
$\proj1$ and $\proj2$. We sometimes leave parameters implicit in dependent
function types, e.g.\ write $\id : A \ra A$ instead of $\id : (A : \Set_i)\ra A
\ra A$. We also use subscripts as a field projection notation for iterated
pairs. For example, for $t : (A : \Set_i) \times (B : \Set_i) \times (f : A \ra
B)$, we use $B_t$ to denote the projection of the second component. Sometimes we
omit the subscript if it is clear from context. When we write ``exists'' in this
paper, we always mean chosen structure given by a $\Sigma$-type.

We write propositional equality as $t = u$, with $\refl_t$ for reflexivity. We
have equality reflection and uniqueness of identity proofs (UIP) for
$\blank=\blank$. The unit type is $\top$, with inhabitant $\tt$.

\todo{compactify cumulativity}

\begin{figure}
\begin{mathpar}
  \inferrule*[lab=universe formation]
             {i < j}
             {\Gamma \vdash \Set_i : \Set_{j}}

  \inferrule*[lab=function formation]
             {\Gamma \vdash A : \U_i \\ \Gamma,\,x : A \vdash B : \U_j}
             {\Gamma \vdash (x : A)\ra B : \U_{\max(i,j)} }

  \inferrule*[lab=$\Sigma$ formation]
             {\Gamma \vdash A : \U_i \\ \Gamma,\,x : A \vdash B : \U_j}
             {\Gamma \vdash (x : A)\times B : \U_{\max(i,j)} }

  \inferrule*[lab=unit formation]
             {\\}
             {\Gamma \vdash \top : \Set_i}

  \inferrule*[lab=equality formation]
             {\Gamma \vdash A : \Set_i \\ \Gamma \vdash t : A \\ \Gamma \vdash u : A}
             {\Gamma \vdash t = u : \Set_i}

  \inferrule*[lab=type lifting]
             {\Gamma \vdash A : \Set_i \\ i \leq j}
             {\Gamma \vdash\,\Lift_j A : \Set_{j}}

  \inferrule*[lab=lift composition]{}{\Lift_j (\Lift_i A) \equiv\,\,\Lift_{j} A}

  \inferrule*[lab=identity lift]{\Gamma \vdash A : \Set_i}{(\Lift_i A) \equiv\,A}

  \inferrule*[lab=term lifting]
             {}
             {\{t\,|\,\Gamma \vdash t : A\} \equiv \{t\,|\,\Gamma \vdash t :\,\,\Lift_i A\}}

  \inferrule*[lab=context lifting]
             {}
             {(\Gamma,\,x : A) \equiv (\Gamma,\,x :\,\,\Lift_i A)}

  \inferrule*[lab=universe lifting]{}{\Lift_j \Set_i \equiv \Set_i}

  \inferrule*[lab=function lifting]
             {}
             {\Lift_i((x : A)\ra B) \equiv (x :\,\Lift_i A)\ra\,\Lift_i B}

  \inferrule*[lab=pair lifting]
             {}
             {\Lift_i((x : A)\times B) \equiv (x :\,\Lift_i A)\,\times\Lift_i B}

  \inferrule*[lab=unit lifting]
             {}
             {\Lift_i\top \equiv \top}

  \inferrule*[lab=equality lifting]
             {}
             {\Lift_i(t = u) \equiv (t = u)}
\end{mathpar}

\caption{Some of the rules for lifting.}
\label{lifting}
\end{figure}

We also have a lifting operation on types, which introduces cumulativity. In
Figure \ref{lifting}, we include an excerpt of cETT's rules. We change
presentation slightly from Sterling: we use a $\U_{\max(i,j)}$ return type in
functions and pairs, instead of having both $A$ and $B$ types in the same
$\Set_i$, and we define type lifting with $i \leq j$ instead of $i < j$. These
changed rules are all derivable from the original ones. In general, we can
expect that lifting never impedes constructions, because it appropriately
computes out of the way.

Of special note is the \emph{term lifting rule}. It is a sort equation, an
equation between sets of terms, expressing that lifted types have exactly the
same terms as unlifted ones. This allows us to have $t :\,\,\Lift_j A$ whenever $t
: A$. Together with the universe lifting rule, this implies $A : \Set_{i+j}$
whenever $A : \Set_i$. Similarly, the lifting rules for functions and pairs give
us cumulativity for $NatAlg$. We derive a notion of subtyping:

\begin{definition}[Cumulative subtyping]
  For $A : \Set_i$, $i \leq j$ and $B : \Set_{j}$, we define $A \leq B : \Set_{j}$
  as $B =\,\,\Lift_j A$.  It follows from term lifting and equality reflection
  that whenever $A \leq B$ and $t : A$, then also $t : B$.
\end{definition}

\subsection{Universe Polymorphism}
\label{sec:universe_polymorphism}

For the sake of simplicity, we do not add this to cETT directly. Instead, we
quantify over levels in an unspecified metatheory \emph{outside} cETT. Hence, a
universe polymorphic cETT term is understood as a $\mathbb{N}$-indexed family of
terms. We reuse the notation of cETT functions for universe polymorphism, e.g.\ as
in the following function:
\[
\lambda\,i.\,\Set_i : (i : \mathbb{N})\ra \Set_{i+1}
\]
In this paper, we do not need to internalize level-polymorphic constructions,
so this setup is sufficient.

@@

\begin{definition}[Notion of model for the theory of signatures]\label{def:tos}
For levels $i$ and $j$, $\ToS_{i,j} : \Set_{\max(i,\,j)+1}$ is a cETT type whose
elements are ToS models (or ToS-algebras). $\ToS_{i,j}$ is an iterated
$\Sigma$-type, containing all of the following components.

@@

\begin{definition}[Notion of induction in an flCwF]\label{def:induction}
Given $\bGamma : \flCwF_i$, we have the following predicate on contexts:
\begin{alignat*}{3}
  & Inductive : \Con_{\bGamma} \ra \Set_i \\
  & Inductive\,\Gamma :\equiv (A : \Ty_{\bGamma}\,\Gamma)\ra \Tm_{\bGamma}\,\Gamma\,A
\end{alignat*}
\end{definition}

@@

In this section we construct QIITs from initial ToS-algebras. From now on, we assume
the existence of such algebras.

@@

\textbf{For each $i$, we assume that $\tos_i : \ToS_{i, 1+i}$ exists}. The reason for the
$1 + i$ level of $\ToS$ sorts is that $\tos_i$ may contain $A : \Set_i$ types in
$\Ty$, and since we intend $\tos_i$ to be a freely generated algebra, the sorts
must be large enough to avoid size paradoxes.

\begin{lemma}[Cumulativity of $\ToS$] If $i \leq i'$ and $j \leq j'$,
  then $\ToS_{i,j} \leq \ToS_{i', j'}$.
\end{lemma}
\begin{proof} $\ToS_{i,j}$ is uniformly defined in $i$, $j$, hence cumulative subtyping extends
  through dependent pairs and functions to $\ToS$.
\end{proof}

@@

>>>>

\input{abbrevs.tex}
\newcommand{\Lift}{\Uparrow}
\newcommand{\ToS}{\mathsf{ToS}}
\newcommand{\ext}{\triangleright}
\newcommand{\emptycon}{\scaleobj{.75}\bullet}
\newcommand{\Pii}{\Pi}
\newcommand{\appi}{\mathsf{app}}
\newcommand{\lami}{\mathsf{lam}}
\newcommand{\Pie}{\Pi^{\mathsf{ext}}}
\newcommand{\appe}{\mathsf{app^{ext}}}
\newcommand{\lame}{\mathsf{lam^{ext}}}
\newcommand{\Piinf}{\Pi^{\mathsf{inf}}}
\newcommand{\appinf}{\mathsf{app^{inf}}}
\newcommand{\laminf}{\mathsf{lam^{inf}}}
\newcommand{\appitt}{\mathop{{\scriptstyle @}}}
\newcommand{\Refl}{\mathsf{Refl}}
\newcommand{\IdU}{\mathsf{IdU}}
\newcommand{\ReflU}{\mathsf{ReflU}}
\newcommand{\Sig}{\mathsf{Sig}}
\newcommand{\ToSSig}{\mathsf{ToSSig}}
\newcommand{\Inductive}{\mathsf{Inductive}}

@@

\todo{revise metatheory intro, use Timany cumulativity}
\todo{recheck universe levels, prettify levels}
\todo{explain the limitations and design choices in univ levels}
\todo{write the rest}
\todo{see if we can cut back on flCwF, add more examples instead}

@@

In Section \ref{sec:metatheory}, we describe the metatheory used in the rest of
the paper.  In Section \ref{sec:tos}, we introduce the theory of QIIT
signatures. TODO

@@

%% We use extensional type theory as metatheory, extended with two features:
%% cumulative universes and a limited form of universe polymorphism.

%% The extensions are used to conveniently talk about large elimination and
%% algebras at different universe levels. For example, consider natural number
%% algebras at level $i$, given as the $\Sigma$-type $NatAlg_i :\equiv(Nat :
%% \Set_i)\times Nat\times(Nat \ra Nat)$. The initial such algebra is the set of
%% natural numbers, which is at level $0$, but in type theory we often want to
%% eliminate into larger $\Set_i$, for example when computing a $Nat$-indexed
%% family of types. Hence, we would like to

%% Cumulativity is used to reduce bureaucratic overhead of
%% handling algebras at different universe levels. For a simple example, consider
%% natural number algebras at level $i$, given as the $\Sigma$-type $NatAlg_i
%% :\equiv(Nat : \Set_i)\times Nat\times(Nat \ra Nat)$. The initial such algebra is
%% the set of natural numbers, which is at level $0$, but in type theory we often
%% want to eliminate into larger $\Set_i$, for example when computing a
%% $Nat$-indexed family of types.

%% We are working in a variant of extensional type theory with cumulative
%% universes, abbreviated as cETT.

%% In this paper we consider QII algebras at arbitrary finite levels, along with
%% large eliminations, where the initial algebra can be at a different level than
%% the target algebra. For a simple example, consider natural number algebras at
%% level $i$, given as the $\Sigma$-type $NatAlg_i :\equiv(Nat : \Set_i)\times
%% Nat\times(Nat \ra Nat)$. The initial such algebra is the set of natural numbers,
%% which is at level $0$, but in type theory we often want to eliminate into larger
%% $\Set_i$, for example when computing a $Nat$-indexed family of types. To support
%% convenient reasoning about levels in this paper, we need two features:
%% cumulativity and a way to quantify over finite levels. The former is essential
%% to reduce bureaucratic overhead, while the latter is required to talk about
%% arbitrary levels.

%% \subsection{Cumulativity}

%% %% Cumulativity makes handling of algebras and constructions at different levels
%% %% much easier. For example, given an algebra $\gamma : NatAlg_i$, cumulativity
%% %% allows us to also have $\gamma : NatAlg_{i+j}$. In a non-cumulative setting such
%% %% as Agda, we may only have for each type $A : \Set_i$ a type $\mathsf{Lift}_{i\,j}\,A :
%% %% \Set_{i+j}$ which is isomorphic to $A$ up to definitional equality.

%% In its most basic form, cumulativity requires that whenever $A : \Set_i$, and $i
%% < j$ then also $A : \Set_j$. We also need cumulativity for $\Sigma$ and function
%% types, in order to have cumulativity for QII algebras in general. For example,
%% we want to have $\gamma : NatAlg_j$ whenever $\gamma : NatAlg_i$ and $i < j$.

%% We use Sterling's cumulative algebraic type theory \cite{sterling2019algebraic} as basis for our metatheory. \todo{REASONS, revamp}

%% %% The reason is that it is the only rigorous
%% %% algebraic definition of cumulativity that we are aware of, and it also

%% %% The reason for this is twofold. First, it
%% %% supports the right kind of cumulativity for our purposes. Second, it is itself a
%% %% finitary QIIT with sort equations, hence this paper provides a model theory for
%% %% it

%% We extend the base theory with an extensional identity type, $\Sigma$-types and
%% the unit type. In \cite{sterling2019algebraic} a proof of canonicity is provided for the base
%% theory, which also includes a standard set-theoretic model. It is
%% straightforward to extend the canonicity proof to cover our additional type
%% formers.

%% From now on, we refer to this theory as cETT (cumulative extensional type
%% theory). When working in cETT, we use the following notation. We have
%% Russell-style universes $\Set_i$ indexed by natural numbers, dependent functions
%% as $(x : A)\ra B$, and dependent pairs as $(x : A)\times B$ with projections
%% $\proj1$ and $\proj2$. We sometimes leave parameters implicit in dependent
%% function types, e.g.\ write $\id : A \ra A$ instead of $\id : (A : \Set_i)\ra A
%% \ra A$. We also use subscripts as a field projection notation for iterated
%% pairs. For example, for $t : (A : \Set_i) \times (B : \Set_i) \times (f : A \ra
%% B)$, we use $B_t$ to denote the projection of the second component. Sometimes we
%% omit the subscript if it is clear from context. When we write ``exists'' in this
%% paper, we always mean chosen structure given by a $\Sigma$-type.

%% We write propositional equality as $t = u$, with $\refl_t$ for reflexivity. We
%% have equality reflection and uniqueness of identity proofs (UIP) for
%% $\blank=\blank$. The unit type is $\top$, with inhabitant $\tt$.

%% \todo{compactify cumulativity}

%% \begin{figure}
%% \begin{mathpar}
%%   \inferrule*[lab=universe formation]
%%              {i < j}
%%              {\Gamma \vdash \Set_i : \Set_{j}}

%%   \inferrule*[lab=function formation]
%%              {\Gamma \vdash A : \U_i \\ \Gamma,\,x : A \vdash B : \U_j}
%%              {\Gamma \vdash (x : A)\ra B : \U_{\max(i,j)} }

%%   \inferrule*[lab=$\Sigma$ formation]
%%              {\Gamma \vdash A : \U_i \\ \Gamma,\,x : A \vdash B : \U_j}
%%              {\Gamma \vdash (x : A)\times B : \U_{\max(i,j)} }

%%   \inferrule*[lab=unit formation]
%%              {\\}
%%              {\Gamma \vdash \top : \Set_i}

%%   \inferrule*[lab=equality formation]
%%              {\Gamma \vdash A : \Set_i \\ \Gamma \vdash t : A \\ \Gamma \vdash u : A}
%%              {\Gamma \vdash t = u : \Set_i}

%%   \inferrule*[lab=type lifting]
%%              {\Gamma \vdash A : \Set_i \\ i \leq j}
%%              {\Gamma \vdash\,\Lift_j A : \Set_{j}}

%%   \inferrule*[lab=lift composition]{}{\Lift_j (\Lift_i A) \equiv\,\,\Lift_{j} A}

%%   \inferrule*[lab=identity lift]{\Gamma \vdash A : \Set_i}{(\Lift_i A) \equiv\,A}

%%   \inferrule*[lab=term lifting]
%%              {}
%%              {\{t\,|\,\Gamma \vdash t : A\} \equiv \{t\,|\,\Gamma \vdash t :\,\,\Lift_i A\}}

%%   \inferrule*[lab=context lifting]
%%              {}
%%              {(\Gamma,\,x : A) \equiv (\Gamma,\,x :\,\,\Lift_i A)}

%%   \inferrule*[lab=universe lifting]{}{\Lift_j \Set_i \equiv \Set_i}

%%   \inferrule*[lab=function lifting]
%%              {}
%%              {\Lift_i((x : A)\ra B) \equiv (x :\,\Lift_i A)\ra\,\Lift_i B}

%%   \inferrule*[lab=pair lifting]
%%              {}
%%              {\Lift_i((x : A)\times B) \equiv (x :\,\Lift_i A)\,\times\Lift_i B}

%%   \inferrule*[lab=unit lifting]
%%              {}
%%              {\Lift_i\top \equiv \top}

%%   \inferrule*[lab=equality lifting]
%%              {}
%%              {\Lift_i(t = u) \equiv (t = u)}
%% \end{mathpar}

%% \caption{Some of the rules for lifting.}
%% \label{lifting}
%% \end{figure}

%% We also have a lifting operation on types, which introduces cumulativity. In
%% Figure \ref{lifting}, we include an excerpt of cETT's rules. We change
%% presentation slightly from Sterling: we use a $\U_{\max(i,j)}$ return type in
%% functions and pairs, instead of having both $A$ and $B$ types in the same
%% $\Set_i$, and we define type lifting with $i \leq j$ instead of $i < j$. These
%% changed rules are all derivable from the original ones. In general, we can
%% expect that lifting never impedes constructions, because it appropriately
%% computes out of the way.

%% Of special note is the \emph{term lifting rule}. It is a sort equation, an
%% equation between sets of terms, expressing that lifted types have exactly the
%% same terms as unlifted ones. This allows us to have $t :\,\,\Lift_j A$ whenever $t
%% : A$. Together with the universe lifting rule, this implies $A : \Set_{i+j}$
%% whenever $A : \Set_i$. Similarly, the lifting rules for functions and pairs give
%% us cumulativity for $NatAlg$. We derive a notion of subtyping:

%% \begin{definition}[Cumulative subtyping]
%%   For $A : \Set_i$, $i \leq j$ and $B : \Set_{j}$, we define $A \leq B : \Set_{j}$
%%   as $B =\,\,\Lift_j A$.  It follows from term lifting and equality reflection
%%   that whenever $A \leq B$ and $t : A$, then also $t : B$.
%% \end{definition}

%% \subsection{Universe Polymorphism}
%% \label{sec:universe_polymorphism}

%% For the sake of simplicity, we do not add this to cETT directly. Instead, we
%% quantify over levels in an unspecified metatheory \emph{outside} cETT. Hence, a
%% universe polymorphic cETT term is understood as a $\mathbb{N}$-indexed family of
%% terms. We reuse the notation of cETT functions for universe polymorphism, e.g.\ as
%% in the following function:
%% \[
%% \lambda\,i.\,\Set_i : (i : \mathbb{N})\ra \Set_{i+1}
%% \]
%% In this paper, we do not need to internalize level-polymorphic constructions,
%% so this setup is sufficient.

@@

\begin{definition}[Notion of model for the theory of signatures]\label{def:tos}
For levels $i$ and $j$, $\ToS_{i,j} : \Set_{\max(i,\,1+j)}$ is a cETT type whose
elements are ToS models (or ToS-algebras). $\ToS_{i,j}$ is an iterated
$\Sigma$-type, containing all of the following components.

@@

\begin{definition}[Notion of induction in an flCwF]\label{def:induction}
Given $\bGamma : \flCwF_i$, we have the following predicate on contexts:
\begin{alignat*}{3}
  & \Inductive : \Con_{\bGamma} \ra \Set_i \\
  & \Inductive\,\Gamma :\equiv (A : \Ty_{\bGamma}\,\Gamma)\ra \Tm_{\bGamma}\,\Gamma\,A
\end{alignat*}
\end{definition}

@@

In this section we construct QIITs from initial ToS-algebras. Before we can assume
the existence of such algebras. First, we need cumulative subtyping for $\ToS$.
\todo{foobar}

\begin{lemma}[Cumulativity of $\ToS$] If $j \leq j'$, then $\ToS_{i,j} \leq \ToS_{i, j'}$.
\end{lemma}
\begin{proof}
  The $j$ universe level only appears in function codomains, hence
  the subtyping rules for functions and $\Sigma$ can be used to get
  $\ToS$ subtyping.
\end{proof}

@@

%% For each level $i$,
%% \begin{enumerate}
%%   \item We assume that $\tos_i : \ToS_{i, 1+i}$ exists.
%%   \item For each $i'$ such that $(1+i) \leq i'$, we assume $\Induction
%% \end{enumerate}

%% \textbf{For each $i$, we assume that $\tos_i : \ToS_{i, 1+i}$ exists}. The reason for the
%% $1 + i$ level of $\ToS$ sorts is that $\tos_i$ may contain $A : \Set_i$ types in
%% $\Ty$, and since we intend $\tos_i$ to be a freely generated algebra, the sorts
%% must be large enough to avoid size paradoxes.

%% \textbf{For each $i$ and $j$, we assume $\mathsf{ToSind :

@@

%% B : Set j

%% (A : Set_i) -> B

%% A : Seti
%% B : Set j
%% (x : A) -> B : Set(max i j)

%% (Con : Setj)x
%% (Ty  : Con -> Setj)x

%% (Pie : (A : Seti) -> (A -> Ty Gamma) -> Ty Gamma
%% : Set(max(suci, j))

%% : Set(max(sucj, max(suci,j)))
%%    max(1+j, max(1+i, j))
%%  = max(1+i, 1+j)

%% REVISE LEVELS!!

%% syntax levels:

  %% ToS(i, 1+i) : Set(max(1+i, 2+i))

%% REWRITE metatheory

%% write about limitations!
